Fly away Peter, come back Paul: the conservation and losses of avian non-coding RNAs

### 2014-03-25 paper/manuscript.tex
Added figs to the manuscript.
Updated RNA-seq table for manuscript.

\author{
Paul P.\ Gardner\correspondingauthor$^{1,2}$
\email{Paul P.\ Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
Sarah W.\ Burge$^3$
\email{swb@ebi.ac.uk},
Maria Ninova$^4$
\email{Maria.Ninova@postgrad.manchester.ac.uk},
Jana Hertel$^5$
\email{Jana Hertel\correspondingauthor - jana@bioinf.uni-leipzig.de},
Stephanie Kehr$^5$
\email{steffi@bierdepot.bioinf.uni-leipzig.de},
Mario Fasold$^5$
\email{mario@bierdepot.bioinf.uni-leipzig.de},
Tammy E.\ Steeves$^1$
\email{tammy.steeves@canterbury.ac.nz},
Sam Griffiths-Jones$^4$
\email{sam.griffiths-jones@manchester.ac.uk}
and
Peter F.\ Stadler\correspondingauthor$^5$
\email{Peter Stadler\correspondingauthor - studla@bioinf.uni-leipzig.de}
}
\address{
\iid(1) School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(2) Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(3) European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK.
\iid(4) Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom.
\iid(5) Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, H{\"a}rtelstrasse 16-18, D-04107 Leipzig, Germany
}

@@

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNAs in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  We show a significant
number of lncRNAs are surprisingly well conserved between birds and
mammals including several intriguing cases where the reported
mammalian lncRNA function is not conserved in birds.  We also
demonstrate extensive conservation of classical ncRNAs (e.g., tRNAs)
and more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in
birds. Furthermore, we have discovered apparent ``losses'' in several
RNA families, these include the divergence of some classical ncRNAs
and the loss of several snoRNAs and microRNAs.  These combined results
illustrate the utility of applying homology based methods for
annotating vertebrate genomes and illustrate many complex evolutionary
patterns within the avian ncRNA cohort.
\end{abstract}

@@

{\bf Say something about the promise of RNA-seq... but requires many
  tissues, developmental stages, fractions(e.g. poly-A, small, ...)
  and the FPR is unknown.}

Despite
promising techniques such as RNA-seq \cite{Croucher:2010}, homology
based methods, namely covariance models (CMs) remain state of the art
for ncRNA analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families. These are used to train
probabilistic models for each family. These models can be used to
generate sequences with similar properties, score the likelihood that
a sequence is generated by the same evolutionary processes as the
training sequences and to build alignments based upon sequence and
structural information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The tRNAscan-SE software package uses CMs to accurately predict
transfer RNAs \cite{Lowe:1997,Chan:2009}.  The Rfam database contains
thousands of curated alignments and consensus structures for diverse
classes of ncRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches dramatically out-perform alternative methods
\cite{Freyhult:2007}, although its sensitivity is limited for the most
rapidly evolving families such as vault RNAs or telomerase RNA
\cite{Menzel:09a}.

{\bf REDUCE?}
The CM based approach works well for almost all classes of ncRNA, but
the long non-coding RNAs (lncRNAs) are a particular challenge
\cite{Guttman:2009}. Recent technological advances have led to
dramatic speed and memory-usage enhancements for CM analyses
\cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. However, CMs
cannot model the exon-intron structures of spliced lncRNAs, nor can
they deal simply with the repeats that many lncRNAs host. Consequently
in the latest release of Rfam the lncRNA families that were added were
composed of local conserved (and possibly structured elements) within
lncRNAs, analogous to the ``domains'' housed within protein sequences
\cite{Burge:2013}. The functions determined to date for lncRNAs range
from regulating chromatin status to chromosomal inactivation
\cite{Rinn:2007,Chow:2005}. Yet functional characterisation of these
genes is a lengthy and expensive process \cite{Guttman:2009}.

The publication of 48 avian genomes, including the previously
published chicken
\cite{International_Chicken_Genome_Sequencing_Consortium:2004}, zebra
finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the newly
published 45 avian genomes \cite{birds:14,birds:14a}, provides
exciting opportunities to explore ncRNA conservation in
unprecedented detail. {\bf MORE ON BIRD GENOMES. ESP. MACRO \& MICRO
  CHROMOSOMES.}

In the following we explore the conservation patterns of the major
classes of avian ncRNA in further detail. The collection of ncRNA
sequences is generally biased towards model organisms
\cite{Gardner:2010,Hoeppner:2012}.  {\bf Using accurate homology
  search tools and evolutionary constraints, we produce an accurate
  set of genome annotations for 48 bird species for ncRNAs that are
  conserved across the avian species. This conservative set of
  annotations is expected to contain the core avian ncRNAs.}
We focus our
report on the unusual results within the avian lineages. These are
either unexpectedly well-conserved RNAs or unexpectedly
poorly-conserved RNAs. The former are RNAs we would not have expected
to be conserved between the birds and the organisms these genes were
initially identified in; Usually, this hypothesis is based upon the
function of the RNA which is not conserved in avian species. The
latter are apparent losses of RNA genes that were expected to be
conserved; usually, this hypothesis is based upon the conservation of
these RNAs in other vertebrate species.  {\bf REWRITE: We use three
  models of ``loss'' that explain the data: Firstly, these could be
  genuine cases of an ancestral gene-loss along the avian
  lineage. Secondly, this could be a case of ``divergence'' where a
  RNA gene has undergone significant sequence and structural
  alterations, so much so, that our homology search tools can no
  longer detect a relationship between vertebrate exemplars and the
  avian varieties. Thirdly, we consider the possibility that the
  available genome assemblies have independently failed to capture
  these genes.}

@@

@@

%% \begin{figure}[ht]
%%   \includegraphics[width=0.45\textwidth]{figures/unusual-conserved.pdf}
%%   \includegraphics[width=0.45\textwidth]{figures/diverged.pdf}
%%   \includegraphics[width=0.1\textwidth]{figures/key2.pdf}
%%   \caption[]{Heatmaps showing the prescence/abscence and approximate
%%     genomic copy number of ``unusually, well conserved RNAs''
%%     (particularly the lncRNAs) on the left and families that have been
%%     identified as surprising RNA Losses, divergence or missing
%%     data. In several cases functionally related families have also
%%     been included, e.g. the RNA components of the major and minor
%%     spliceosomes: U1, U2, U4, U5 and U6; and U11, U12, U4atac, U5 and
%%     U6atac, respectively.}\label{fig:1}
%% \end{figure}

@@

{\bf TABLE ``AVERAGE'' NUMBER OF ncRNAs IN A BIRD GENOME?}

%% \begin{figure}[ht]
%%   \includegraphics[width=0.75\textwidth]{figures/More-cand.pdf}
%%   \caption[]{Additional homologs of some sparsely represented RNA
%%     families were discovered using dedicated search strategies
%%     combined with highly sensitive settings, synteny information,
%%     lineage-specific CMs and subsequent manual
%%     inspection.}\label{fig:2}
%% \end{figure}

>>>>

\author{
Paul P.\ Gardner\correspondingauthor$^{1,2}$
\email{Paul P.\ Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
Mario Fasold$^5$
\email{mario@bierdepot.bioinf.uni-leipzig.de},
Sarah W.\ Burge$^3$
\email{swb@ebi.ac.uk},
Maria Ninova$^4$
\email{Maria.Ninova@postgrad.manchester.ac.uk},
Jana Hertel$^5$
\email{Jana Hertel\correspondingauthor - jana@bioinf.uni-leipzig.de},
Stephanie Kehr$^5$
\email{steffi@bierdepot.bioinf.uni-leipzig.de},
Tammy E.\ Steeves$^1$
\email{tammy.steeves@canterbury.ac.nz},
Sam Griffiths-Jones$^4$
\email{sam.griffiths-jones@manchester.ac.uk}
and
Peter F.\ Stadler\correspondingauthor$^5$
\email{Peter Stadler\correspondingauthor - studla@bioinf.uni-leipzig.de}
}
\address{
\iid(1) School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(2) Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(3) European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK.
\iid(4) Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom.
\iid(5) Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, H{\"a}rtelstrasse 16-18, D-04107 Leipzig, Germany
}

@@

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNAs in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase.  We show a significant
number of lncRNAs are surprisingly well conserved between birds and
mammals including several intriguing cases where the reported
mammalian lncRNA function is not conserved in birds.  We also
demonstrate extensive conservation of classical ncRNAs (e.g., tRNAs)
and more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in
birds. Furthermore, we have discovered apparent ``losses'' in several
RNA families, these illustrate the complexity of bird genomes. In
particular, issues with assembling microchromosomes using
next-generation sequencing methods. These combined results illustrate
the utility of applying homology based methods for annotating
vertebrate genomes and illustrate many complex evolutionary patterns
within the avian ncRNA cohort.
\end{abstract}

@@

{\bf needs work:} New sequencing technologies have dramatically
expanded the rate at which ncRNAs are discovered and their functions
are determined \cite{}. However, in order to determine the full range
of ncRNAs across multiple species we would require multiple RNA
fractions (e.g. long and short), in multiple species, in multiple
developmental stages and tissues types. The costs of this approach are
still prohibitive. Consequently, for this work we have concentrated on
bioinformatic approaches. Primarily, homology based methods, namely
covariance models (CMs). These remain state of the art for ncRNA
analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009} and have well
established sensitivity and specificity rates \cite{Freyhult:2007}.
The CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families. These are used to train
probabilistic models for each family. These models can be used to
generate sequences with similar properties, score the likelihood that
a sequence is generated by the same evolutionary processes as the
training sequences and to build alignments based upon sequence and
structural information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The tRNAscan-SE software package uses CMs to accurately predict
transfer RNAs \cite{Lowe:1997,Chan:2009}.  The Rfam database contains
thousands of curated alignments and consensus structures for diverse
classes of ncRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches dramatically out-perform alternative methods
\cite{Freyhult:2007}, although its sensitivity is limited for the most
rapidly evolving families such as vault RNAs or telomerase RNA
\cite{Menzel:09a}.

{\bf REDUCE? too much talking about what it can't to compared to what it can do...}
The CM based approach works well for almost all classes of ncRNA, but
the long non-coding RNAs (lncRNAs) are a particular challenge
\cite{Guttman:2009}. Recent technological advances have led to
dramatic speed and memory-usage enhancements for CM analyses
\cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. However, CMs
cannot model the exon-intron structures of spliced lncRNAs, nor can
they deal simply with the repeats that many lncRNAs host. Consequently
in the latest release of Rfam the lncRNA families that were added were
composed of local conserved (and possibly structured elements) within
lncRNAs, analogous to the ``domains'' housed within protein sequences
\cite{Burge:2013}. The functions determined to date for lncRNAs range
from regulating chromatin status to chromosomal inactivation
\cite{Rinn:2007,Chow:2005}. Yet functional characterisation of these
genes is a lengthy and expensive process \cite{Guttman:2009}.

The publication of 48 avian genomes, including the previously
published chicken
\cite{International_Chicken_Genome_Sequencing_Consortium:2004}, zebra
finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the recently
published 45 avian genomes \cite{birds:14,birds:14a,Huang:2013,Zhan:2013,Shapiro:2013,Howard:2013,Li:2014}, provides
exciting opportunities to explore ncRNA conservation in
unprecedented detail. 
%{\bf MORE ON BIRD GENOMES. ESP. MACRO \& MICRO
%  CHROMOSOMES.}

In the following we explore the conservation patterns of the major
classes of avian ncRNA in further detail. The collection of ncRNA
sequences is generally biased towards model organisms
\cite{Gardner:2010,Hoeppner:2012}.  {\bf Using accurate homology
  search tools and evolutionary constraints, we produce an accurate
  set of genome annotations for 48 bird species for ncRNAs that are
  conserved across the avian species. This conservative set of
  annotations is expected to contain the core avian ncRNAs.}
We focus our
report on the unusual results within the avian lineages. These are
either unexpectedly well-conserved RNAs or unexpectedly
poorly-conserved RNAs. The former are RNAs we would not have expected
to be conserved between the birds and the organisms these genes were
initially identified in; Usually, this hypothesis is based upon the
function of the RNA which is not conserved in avian species. The
latter are apparent losses of RNA genes that were expected to be
conserved; usually, this hypothesis is based upon the conservation of
these RNAs in other vertebrate species.

Here, we consider three categories of apparent loss: First, genuine
gene losses in the avian lineage. Second, ``divergence'' where RNA genes
have undergone such significant sequence and structural alternations
that homology search tools can no longer detect a relationship between
other vertebrate exemplars and avian varieties. Third, ``missing'' genes
that failed to be captured in the available, largely
fragmented, avian genomes. We postulate that the latter
category is likely to be prevalent in comparative avian genome studies
given the distinctive organisation of the avian genome. Namely, the
avian karyotype is characterised by a large number of chromosomes
(average $2n \approx 80$) generally consisting of a few larger
``macrochromosomes'' and many smaller ``microchromosomes'' \cite{Griffin:2007}. This 'so many, so small' pattern presents significant
assembly challenges \cite{Ellegren:2005}. Indeed, of the 48 published avian
genomes, 20 of which are high-coverage ($>50X$), only two are
chromosomally assembled (chicken and zebra finch; \cite{birds:14}).

%% {\bf REWRITE: We use three
%%   models of ``loss'' that explain the data: Firstly, these could be
%%   genuine cases of an ancestral gene-loss along the avian
%%   lineage. Secondly, this could be a case of ``divergence'' where a
%%   RNA gene has undergone significant sequence and structural
%%   alterations, so much so, that our homology search tools can no
%%   longer detect a relationship between vertebrate exemplars and the
%%   avian varieties. Thirdly, we consider the possibility that the
%%   available genome assemblies have independently failed to capture
%%   these genes.}

@@

We thank Matt Schwatz (Harvard) and Igor Ulitsky (Weizmann Institute
of Science) for access to the RNA-seq data.

@@

 \begin{figure}[ht]
   \includegraphics[width=0.5\textwidth]{figures/figure1.pdf}
 \end{figure}

@@

 \begin{figure}[ht]
   \includegraphics[width=0.5\textwidth]{figures/figure2.pdf}
 \end{figure}

  \subsection*{Table 1 - A summary of ncRNA genes in human, chicken and all bird genomes}

This table contains the total number of annotated ncRNAs from
different RNA types in human, chicken and the median number for each
of the 48 birds. The number of chicken ncRNA that show evidence for
expression is also indicated (the percentage is given in parentheses).

False positive rate less the 10\%:

\begin{tabular}{|r|r|r|r|l|}
\hline 
\multicolumn{5}{|l|}{{\bf ncRNA genes in human, chicken and all bird genomes}}\\
\hline 
                &                  &                   & \multicolumn{1}{|l|}{Chicken ncRNAs } & \\
                &                  &                   & \multicolumn{1}{|l|}{confirmed with RNA-seq} & \\
\hline
Number in human & median(48 birds) & Number in chicken & max(RNA$_i$)$>13.0$ & RNA type\\
\hline
62&25.0&34&12 (35.3\%) &Long non-coding RNA\\ 
356&499.5&427&272 (63.7\%) &microRNA\\ 
281&120.0&106&88 (83.0\%) &C/D box snoRNA\\ 
336&85.5&68&46 (67.6\%) &H/ACA box snoRNA\\ 
34&13.0&12&11 (91.7\%) &Small cajal body RNA\\ 
1754&48.5&71&29 (40.8\%) &Major spliceosomal RNA\\ 
58&3.0&6&2 (33.3\%) &Minor spliceosomal RNA\\ 
525&82.0&122&79 (64.8\%) &Cis-regulatory element\\ 
316&6.5&9&3 (33.3\%) &7SK RNA\\ 
1&0.0&2&0 (0.0\%) &Telomerase RNA\\ 
9&0.0&2&1 (50.0\%) &Vault RNA\\ 
892&3.0&3&2 (66.7\%) &Y RNA\\ 
1084&173.5&300&256 (85.3\%) &Transfer RNA\\ 
80&9.5&4&2 (50.0\%) &Transfer RNA pseudogene\\ 
941&3.0&4&2 (50.0\%) &SRP RNA\\ 
607&7.0&22&10 (45.5\%) &Ribosomal RNA\\ 
4&1.0&2&2 (100.0\%) &RNase P/MRP RNA\\ 
\hline
7340&1080.0&1194&817 (68.4\%) &Total\\ 
\hline
\end{tabular}
